CS6110: Software Verification Final Project: Formal Verification of Rust Programs with Creusot

### 2023-05-05 paper.tex
Writeup section on using different proof dischargers with Why3

\documentclass[12pt]{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{biblatex}
\usepackage{hyperref}
\usepackage{fullpage}
\usepackage{listings}
\usepackage{xcolor}

@@

Consider a simple verification of a \texttt{sum} function. This is how we would write it with WhyML:

\begin{verbatim}
\end{verbatim}

@@

Why3 enables us to use many different proof automators and assistants at the same time.
We will leave a full explaination of how to use the Why3 IDE to do this to their documentation\footnote{https://why3.lri.fr/doc/starting.html#sec-gui}, but suffice it to say that one of the most interesting abilities of Why3 is to take a Why3 goal and use any sort of prover on it to discharge it.

For example, SMT solvers can discharge proofs automatically.

\subsection{SMT Solvers}

\subsection{Proof automation}

% Brief discussion on what e.g. z3 &co. are

@@

\subsection{Example: verifying a very simple function}

\subsection{Example: a more complex example}

\section{Experience Report: verifying a library} % ← maybe; big maybe

@@

    CreuSAT is a SAT solver verified with Creusot.

@@

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{biblatex}
\usepackage{hyperref}
\usepackage{fullpage}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsfonts}

@@

Consider verifying a simple fact about arithmetic like $\forall x \in \mathbb{Z}, 2x = x + x$.
This is how we would write that in WhyML:

\begin{figure}[h]
\centering
\begin{verbatim}
module SumEquiv
  use int.Int

  goal SimpleThing: forall x:int. 2 * x = x + x
end
\end{verbatim}
\caption{SumEquiv: a very simple theorem showing that $\forall x \in \mathbb{Z}, 2x = x + x$}
\label{fig:sumequiv}
\end{figure}

This is an incredibly simple example; we will leave a full description of what the language can do to the specification\footnote{https://why3.lri.fr/doc/syntaxref.html}, but suffice it to say that WhyML has all of the language constructs that you would expect from a formal logic system.

% TODO: we should highlight some differences

@@

The really exciting part of Why3 is how it can translate WhyML into input to various other theorem provers, interactive and automatic.
For example, we can ask the Alt-Ergo SMT solver to check that figure \ref{fig:sumequiv} holds, which it can do automatically.
There are several SMT solvers (e.g. Z3, CVC4, etc.) that Why3 can convert WhyML into and leverage.

% What are the strengths of one over another?

As nice as SMT solvers are, they are not able to do everything that interactive theorem provers can do.
Interactive provers---like Coq or Idris---allow a human to take control of the proof.
Fortunately, Why3 also supports using interactive provers to discharge individual proof steps.

We can ask Why3 to discharge the proof of \texttt{SimpleThing} with Coq.
Why3 opens the \texttt{coqide} with the following-filled for us:

\begin{verbatim}
(* This file is generated by Why3's Coq driver *)
(* Beware! Only edit allowed sections below    *)
Require Import BuiltIn.
Require BuiltIn.
Require int.Int.

(* Why3 goal *)
Theorem SimpleThing : forall (x:Numbers.BinNums.Z), ((2%Z * x)%Z = (x + x)%Z).
(* Why3 intros x. *)
intros x.

Qed.
\end{verbatim}

Now it is up to us to fill in the rest of the proof.

We have to use the ZArith module from the Coq standard library\footnote{} because Why3 works with integers, not naturals.
We can add this line after the \texttt{intros x.} to get Coq to accept the proof:

\begin{verbatim}
rewrite Z.add_diag. reflexivity.
\end{verbatim}

With that in place, we ask Why3 to ``replay obsolete proofs'' (in Why3 parlance) and it will run Coq, make sure that the intros and the theorem in the source still line up, and accept that theorem as proven if Coq succeeds.

Note that all proving \emph{must} go after the \texttt{(* Why3 intros x. *)} part and before the \texttt{Qed.} line.
This is so that Why3 can parse the source file and reliably rebuild the proof later on.

@@

Creusot works by installing a helper tool to run with Rust's \texttt{cargo} build and package manager.

\subsection{Example: verifying a simple function}

@@

    CreuSAT is a SAT solver verified with Creusot.
    CreuSAT pushes the boundaries of what Creusot can do.
    We were able to verify the source of CreuSAT after many tears and much lost sleep.
    Once we finally got the entire stack to compile, verification went off without a hitch thanks to some rather simple automatic verification strategies provided by CreuSAT.
    This showed off how Why3 can make verifying programs almost entirely automatic.

    CreuSAT shows that Creusot \emph{can} scale to significantly-sized projects.
    The UX and the documentation just aren't there for mainstream use, however.

@@

    We ran out of time before we were able to evaluate Kani for this project.
    It does seem like an intriguing piece of software however.
